Probabilities of polar aligned disks in binary star systems

### 2024-09-17 src/tex/ms.tex
Fix problems and integrate f_polar

To get the probability of a polar orbit given randomly distributed $\Omega$ we can use a modified version of Equation 8 from \citet{ceppi2024}.
\begin{equation}
    P(j,e_{\rm b}, i) = 
    \left \{
    \begin{array}{lr}
         1 - \frac{2}{\pi} \Omega_{\rm min} & \text{if} ~i > i_{\rm crit} \\
        0 & \text{if} ~i \le i_{\rm crit}
    \end{array}
    \right .
\end{equation}
From which the fraction of total polar orbits can be calculated
\begin{equation}
    \label{eq:fp_int}
    f_{\rm polar} = \iiint P(j,e_{\rm b}, i)\, dj\,de_{\rm b}\,di
\end{equation}
This integral is straightforward to compute numerically, and allows for any desired distribution in $j$, $e_{\rm b}$, or $i$, so long as they are normalized. For example, an isotropic distribution of disks would require we replace $di$ with $\frac{1}{2}\sin{u}\,du$ for $u \in [0,\pi]$.

@@

\TJ{Does this explain why we see prograde precession when $i=\pi$ for high $j$. Will prograde precession of an anti-aligned disk lead to alignment?}

\section{Runge-Kutta-Fehlburg method}
\label{sec:rk}
\citet[Equations (7-10)]{martin2019} also provide a set of coupled first-order differential equations that describe the evolution of $e_{\rm b}$ along with the direction of the disk angular momentum $(\ell_x, \ell_y, \ell_z)$, based on previous work by \citet{farago2010}. We solve these equations using the Runge-Kutta-Fehlberg (RKF) method \TJ{ref?} -- a numerical method similar to the common 4th order Runge-Kutta integrator, but that computes a 5th order solution to allow for a variable step size. This method treats the binary in the quadrupole approximation, meaning that it is not sensitive to the effects of the binary mass fraction or orbital dynamics that occur on timescales less than the binary orbital period. However, the computational cost is extremely low; we are able to run $10^4$-$10^5$ simulations per second -- about 10,000 times faster than using \sc{REBOUND} in Section \ref{sec:reb}.

@@

\section{Simulating with \sc{REBOUND}}
\label{sec:reb}

We use the N-body integration software \sc{REBOUND} \citep{rebound} to simulate a circumbinary system with an additional level of sophistication compared to those described in Section \ref{sec:rk}, in which the binary was treated in the quadrupole approximation. We model the disk as a point mass with its same orbital parameters and angular momentum.

@@

\TJ{(How to account for the changing position of the COM was $f_{\rm b}$ changes? This is used to define the coordinate system)}

\begin{figure}
    \begin{centering}
        \includegraphics[width=0.5\textwidth]{figures/compare_massive.pdf}
        \caption{
            MC results (points) for $e_{\rm b}=0.4$, $j=0.67$, and $f_{\rm b}=0.5$ compared to RKF results for the same setup (contour lines). This value of $j$ is chosen here as it is the critical value \citep[see][]{martin2019,abod2022}, beyond which retrograde dynamical states are not allowed.
            Note that some points that \sc{REBOUND} classifies as polar fall outside the RKF polar region.
            The MC result gives $f_{\text{polar}} = \variable{output/mc_out.txt}$ while the RKF result gives $f_{\text{polar}} = \variable{output/rk_out.txt}$ with
            numerical error on the order of $\variable{output/rk_err.txt}$. 
        }
        \script{compare_massive.py}
        \label{fig:rkf}
    \end{centering}
\end{figure}

@@

In Section \ref{sec:analytic} we made the simplifying assumption that $\bm{l}$ is isotropic. Note that this is in contrast to the exponential distribution in
mutual inclination employed by \citet{ceppi2024}. The decaying exponential prefers disks that are aligned with the binary -- valid in the wide-binary regime, but
hydrodynamic simulations by \citep{elsender2023} found that, for binaries with separations on the order of 1 AU, mutual inclinations are seemingly randomly
distributed between $0$ and $\pi$. % See their fig 8

\begin{figure*}
    \begin{centering}
        \includegraphics[width=1\textwidth]{figures/compare_ebin.pdf}
        \caption{
            Polar fraction as a function of $e_{\rm b}$ for various values of $j$. Each fraction is calculated using a $100\times 100$ grid of RKF integrations.
            The dashed lines shows the analytic solution for a massless disk from \citet{zanazzi2018} (in yellow) and from \citet{martin2019} (in gray), integrated numerically.
            There are two competing effects here:
            In the low-$j$ regime the polar fraction is a strong function of $e_{\rm b}$. However, as $j$ increases, $f_{\rm polar}$ is only sensitive to eccentricity
            when $e_{\rm b} \sim 1$, and otherwise prefers a flat distribution.
        }
        \script{compare_ebin.py}
        \label{fig:ebin}
    \end{centering}
\end{figure*}

First, let's not consider the other two distributions, and leave $j$ and $e_{\rm b}$ as free parameters. Figure \ref{fig:ebin} shows that the polar fraction
as a function of $e_{\rm b}$ can be modulated by $j$ to display two behaviors. There exists an analytic solution by \citet{zanazzi2018}, and the tendency is
for $f_{\rm polar}$ to follow this solution either when $j$ is low or $e_{\rm b}$ is very high. Contrarily, if $j$ is sufficiently large, then a flat distribution
(e.i. no dependence on $e_{\rm b}$) is preferred except for when $e_{\rm b}$ is very nearly $1$ (see Section \ref{subsubsec:hi_j}).

\subsection{An upper limit on the polar fraction}
\label{subsec:fp_limit}
While the true astrophysical distributions of $i$, $e_{\rm b}$, and $j$ are unknown, we can still place constraints on the value of $f_{\rm polar}$.
As discussed previously, if the disk has no preference for coplanar alignment then we can replace $di$ in Equation \ref{eq:fp_int} with $\frac{1}{2}\sin{du_i}du_i$, integrating
from 0 to $\pi$.

@@

Differentiating $\gamma$ tells us regardless of $\chi$ that $\frac{d\gamma}{de_{\rm b}} < 0$, so
\begin{equation}
    \frac{d\Omega_{\rm min}}{de_{\rm b}} < 0
\end{equation}
or that the polar fraction always increases with increasing binary eccentricity.
To truly get an upper limit we should choose from a family of distributions one that prefers high eccentricities. However, typical distribution
functions have the form $P(e_{\rm b}) \propto e_{\rm b}^\alpha$ \citep[e.g.][]{hwang2022,ceppi2024}, in which the ``upper limit'' distribution
is one in which $\alpha \rightarrow \infty$ and $P(e_{\rm b})$ approaches the delta function $\delta(e_{\rm b} - 1)$. This would result in a polar frequency
of 1. \citet{hwang2022} suggests that for binaries with separations less than 100 AU it is appropriate to adopt a random distribution (i.e. $\alpha=0$). We then simply
integrate $de_{\rm b}$ from 0 to 1.

\TJ{Now for the question of how to integrate $j$...}

>>>>

To get the probability of a polar orbit given randomly distributed $\Omega$ we can use a modified version of Equation 8 from \citet{ceppi2024}.
\begin{equation}
    P(j,e_{\rm b}, i) = 
    \left \{
    \begin{array}{lr}
         1 - \frac{2}{\pi} \Omega_{\rm min} & \text{if} ~i > i_{\rm crit} \\
        0 & \text{if} ~i \le i_{\rm crit}
    \end{array}
    \right .
\end{equation}
From which the fraction of total polar orbits can be calculated
\begin{equation}
    \label{eq:fp_int}
    f_{\rm polar} = \iiint P(j,e_{\rm b}, i)\,p_j(j)\,p_{e_\text{b}}(e_\text{b})\,p_i(i)\, dj\,de_{\rm b}\,di
\end{equation}
where $p_j(j)$, $p_{e_\text{b}}(e_\text{b})$, and $p_i(i)$ are the distributions of $j$, $e_\text{b}$, and $i$, respectively.

This integral is straightforward to compute numerically, and allows for any desired distribution in $j$, $e_{\rm b}$, or $i$, so long as they are normalized. For example, an isotropic distribution of disks would require $p_i(i) = \frac{1}{2}\sin{i}$ integrating from $0$ to $\pi$.

@@

\section{Runge-Kutta-Fehlburg method}
\label{sec:rk}
\citet[Equations (7-10)]{martin2019} also provide a set of coupled first-order differential equations that describe the evolution of $e_{\rm b}$ along with the direction of the disk angular momentum $(\ell_x, \ell_y, \ell_z)$, based on previous work by \citet{farago2010}. We solve these equations using the Runge-Kutta-Fehlberg (RKF) method \TJ{ref?} -- a numerical method similar to the common 4th order Runge-Kutta integrator, but that computes a 5th order solution to allow for a variable step size. This method treats the binary in the quadrupole approximation, meaning that it is not sensitive to the effects of the binary mass fraction or orbital dynamics that occur on timescales less than the binary orbital period. However, the computational cost is extremely low; we are able to run $10^4$-$10^5$ simulations per second -- about 10,000 times faster than using {\sc REBOUND} in Section \ref{sec:reb}.

@@

\section{Simulating with {\sc REBOUND}}
\label{sec:reb}

We use the N-body integration software {\sc REBOUND} \citep{rebound} to simulate a circumbinary system with an additional level of sophistication compared to those described in Section \ref{sec:rk}, in which the binary was treated in the quadrupole approximation. We model the disk as a point mass with its same orbital parameters and angular momentum.

@@

\begin{figure}
    \begin{centering}
        \includegraphics[width=0.5\textwidth]{figures/compare_massive.pdf}
        \caption{
            MC results (points) for $e_{\rm b}=0.4$, $j=0.67$, and $f_{\rm b}=0.5$ compared to RKF results for the same setup (contour lines). This value of $j$ is chosen here as it is the critical value \citep[see][]{martin2019,abod2022}, beyond which retrograde dynamical states are not allowed.
            Note that some points that {\sc REBOUND} classifies as polar fall outside the RKF polar region.
            The MC result gives $f_{\text{polar}} = \variable{output/mc_out.txt}$ while the RKF result gives $f_{\text{polar}} = \variable{output/rk_out.txt}$ with
            numerical error on the order of $\variable{output/rk_err.txt}$. 
        }
        \script{compare_massive.py}
        \label{fig:rkf}
    \end{centering}
\end{figure}

Figure \ref{fig:rkf} shows that, while the RKF and {\sc rebound} simulations produce similar results, they are not identical. Specifically noticeable are points sampled by the MC algorithm that lie outside the RKF libration region (with a higher inclination), but are found by {\sc rebound} to librate. There are also points with the opposite discrepancy -- those that lie in the RKF libration region but produce prograde precession -- but they are fewer and not as noticeable by eye. \TJ{(We should try to figure out why this happens)}. However, on the population level, the two measurements of $f_\text{polar}$ lie within the MC uncertainty. As the RKF method is much faster computationally, we will use those results in discussing populations.

@@

In Section \ref{sec:analytic} we made the simplifying assumption that $\bm{l}$ is isotropic. Note that this is in contrast to the exponential distribution in
mutual inclination employed by \citet{ceppi2024}. The decaying exponential prefers disks that are aligned with the binary -- valid in the wide-binary regime, but
hydrodynamic simulations by \citet{elsender2023} found that, for binaries with separations on the order of 1 AU, mutual inclinations are seemingly randomly
distributed between $0$ and $\pi$. % See their fig 8

\begin{figure*}[!htbp]
    \begin{centering}
        \includegraphics[width=1\textwidth]{figures/compare_ebin.pdf}
        \caption{
            Polar fraction as a function of $e_{\rm b}$ for various values of $j$. Each fraction is calculated using a $100\times 100$ grid of RKF integrations.
            The dashed lines shows the analytic solution for a massless disk from \citet{zanazzi2018} (in yellow) and from \citet{martin2019} (in gray), integrated numerically.
            There are two competing effects here:
            In the low-$j$ regime the polar fraction is a strong function of $e_{\rm b}$. However, as $j$ increases, $f_{\rm polar}$ is only sensitive to eccentricity
            when $e_{\rm b} \sim 1$, and otherwise prefers a flat distribution.
        }
        \script{compare_ebin.py}
        \label{fig:ebin}
    \end{centering}
\end{figure*}

First, let's not consider the other two distributions, and leave $j$ and $e_{\rm b}$ as free parameters. Figure \ref{fig:ebin} shows that the polar fraction
as a function of $e_{\rm b}$ can be modulated by $j$ to display two behaviors. There exists an analytic solution by \citet{zanazzi2018}, and the tendency is
for $f_{\rm polar}$ to follow this solution either when $j$ is low or $e_{\rm b}$ is very high. Contrarily, if $j$ is sufficiently large, then a flat distribution
(e.i. no dependence on $e_{\rm b}$) is preferred except for when $e_{\rm b}$ is very nearly $1$ (the Kozai-Lidov regime, see Section \ref{subsec:high_j}).

\subsection{An upper limit on the polar fraction}
\label{subsec:fp_limit}
While the true astrophysical distributions of $i$, $e_{\rm b}$, and $j$ are unknown, we can still place constraints on the value of $f_{\rm polar}$.
As mentioned previously, if the disk has no preference for coplanar alignment, then
\begin{equation}
    p_i(i) = \frac{1}{2} \sin{i}
\end{equation}
and we integrate from $0$ to $\pi$.

@@

Differentiating $\gamma$ tells us regardless of $\chi$ that $\frac{d\gamma}{de_{\rm b}} < 0$, so
\begin{equation}
    \frac{d\Omega_{\rm min}}{de_{\rm b}} < 0
\end{equation}
or that the polar fraction always increases with increasing binary eccentricity.

To truly get an upper limit we should choose from a family of distributions one that prefers high eccentricities. However, typical distribution
functions have the form $P(e_{\rm b}) \propto e_{\rm b}^\alpha$ \citep[e.g.][]{hwang2022,ceppi2024}, in which the ``upper limit'' distribution
is one in which $\alpha \rightarrow \infty$ and $P(e_{\rm b})$ approaches the delta function $\delta(e_{\rm b} - 1)$. This would result in a polar frequency
of 1. \citet{hwang2022} suggests that for binaries with separations less than 100 AU it is appropriate to adopt a random distribution (i.e. $\alpha=0$). We then simply
integrate from 0 to 1 with
\begin{equation}
    p_{e_\text{b}}(e_\text{n}) = 1
\end{equation}

We elect not to make any assumptions about the distribution of $j$. We will instead calculate $f_\text{polar}$ for various constant $j$. This is analogous by replacing $j$ in Equation (\ref{eq:fp_int}) with $j'$, and letting $p_{j'}(j') = \delta(j'-j)$. We are now interested in the polar fraction at fixed $j$:
\begin{equation}
    f_\text{polar}(j) = \int_{0}^{\pi} \int_{0}^{1} P(j,e_\text{b},i)\, \frac{1}{2} \sin{i}\,de_\text{b}\,di
\end{equation}

Figure \ref{fig:fpol_j} shows this integral computed for a range of $j$. We see that for low $j$, $f_\text{polar} \lessapprox 0.5$.

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/fpol_j.pdf}
    \caption{
        Polar fraction $f_\text{polar}$ upper limit as a function of normalized disk angular momentum $j$. These simulations assume a population of binary systems with constant $j$, random $e_\text{b}$, and disks whose angular momentum vectors are distributed isotropically with respect to the binary angular momentum. These values were computed using the analytic technique described in Section \ref{sec:analytic}.
    }
    \script{f_p_of_j.py}
    \label{fig:fpol_j}
\end{figure*}
